---
title: "Reproducible Research: Peer Assessment 1"
author: "Rahulg13
output: 
html_document:
    keep_md: true
---


## Loading and preprocessing the data
The file available in the folder is a zip file. So, first we need to unzip it and then import the data into R. This can be done as follows - 

```{r}
unzip("activity.zip")
# Reading the data 
d <- read.csv("activity.csv")
```

Then, for processing the dates as "Date" variable, it needs to be transformed into date format which is done as follows - 
```{r}
# processing data
d$date <- as.Date(d$date)
```


## What is mean total number of steps taken per day?
Here three questions are pertinent. We solve them one-by-one.

### 1. Calculate the total number of steps taken per day

```{r}
datesum <- tapply(d$steps, d$date, sum)
```

### 2.  Make a histogram of the total number of steps taken each day

```{r}
hist(datesum, breaks = 8, xlab = "steps count", main = "Steps taken per day")
```

### 3. Calculate and report the mean and median of the total number of steps taken per day

```{r}
stepmean <- mean(datesum, na.rm = TRUE)
stepmedian <- median(datesum, na.rm = TRUE)
stepmean
stepmedian
```

## What is the average daily activity pattern?
We want to create a time series plot for the step count in each interval.
* First, interval-wise average step count need to be calculated
* Second, all the intervals need to be stored in a variable 
* Third, a data frame needs to be created by bringing above two together 
* Fourth, plotting the required figure
```{r}
intervalcount <- tapply(d$steps, d$interval, mean, na.rm = TRUE)
intervals <- levels(as.factor(d$interval))
q2 <- cbind(intervalcount, intervals)
q2 <- as.data.frame(q2)
q2$intervals <- as.numeric(q2$intervals)
q2$intervalcount <- as.numeric(as.character(q2$intervalcount))
library(ggplot2)
#g <- ggplot(q2, aes(y = intervalcount, x = intervals)) + geom_point()
#g <- g + geom_smooth()
g <- ggplot(q2, aes(y = intervalcount, x = intervals)) + geom_line()
g <- g + ggtitle("Step Count  vs  Time interval")
g
```

Next, we need to find which interval contains the maximum number of steps.
```{r}
#Finding maximum
maxindex <- which.max(q2$intervalcount)
maxinterval <- q2$intervals[maxindex]
maxinterval
```

## Imputing missing values
```{r}
#Finding NAs
naindex <- which(is.na(d$steps))

# Creating new Dataset
d2 <- d

#Replacing NAs with average of that interval
d2$steps[naindex] <- intervalcount[naindex]

# Plotting new histogram
datesum2 <- tapply(d2$steps, d2$date, sum)
hist(datesum2, breaks = 8, xlab = "steps count", main = "Steps taken per day")

#Finding new mean and median
stepmean2 <- mean(datesum, na.rm = TRUE)
stepmedian2 <- median(datesum, na.rm = TRUE)
```

We need to find out the difference between the two. 
```{r}
print("Difference between means")
stepmean - stepmean2
print("Difference between medians")
stepmedian - stepmedian2
```

## Are there differences in activity patterns between weekdays and weekends?
```{r}

library(lubridate)
library(dplyr)
d <- mutate(d, weekend = ((wday(date) == 1) | (wday(date) == 7)))
d$weekend <- as.factor(d$weekend)
levels(d$weekend) <- c("Weekday", "Weekend")
divided <- with(d, tapply(steps, list(interval, weekend), mean, na.rm =  TRUE))
divided <- as.data.frame(divided)
l <- length(divided[,1])
divided <- unlist(divided)
daytype <- c(rep("Weekday", l), rep("Weekend", l))
intervaltype <- c(intervals, intervals)
divided <- cbind(divided, daytype, intervaltype)
divided <- as.data.frame(divided)
names(divided) <- c("steps", "day_type", "interval")
divided$interval <- as.numeric(divided$interval)
divided$steps <- as.numeric(divided$steps)

g2 <- ggplot(divided, aes(x = interval, y = steps))
g2 <- g2 + geom_line(col = "blue") + facet_grid(day_type ~ .)
g2
```